Analyses of 27 years of stand structure development and basal area growth response to a range of initial basal area density levels, 1992-2019

### 2020-05-06 manuscript/EP1162_manuscript_draft.Rmd
manny work

---
title: Analyses of 27 years of stand structure development and basal area growth response to a range of initial basal area density levels, 1992-2019 
author: "Mike Jull and Hardy Griesbauer"
date: "29/03/2020"
output:
  word_document:
    reference_docx: wordTemplate2.docx
  html_document:
    df_print: paged
csl: cjfr.csl
bibliography: ep1162Bib.bib
---

@@

# Load libraries
library(tidyverse)
library(here)
library(lme4)
library(lmerTest)
library(broom)
library(MuMIn)
library(ggeffects)
library(car)
library(flextable)
library(captioner)

@@

@@

### Tree-level growth responses
To test the hypothesis that tree-level growth response over the 1992-2019 period varied with interactions between (i) tree basal area at the time of treatment (1992); (ii) residual basal area; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 basal area increment as the response variable, and treatment, species and log-transformed 1992 basal area as interacting fixed effects, with plot as a random intercept effect.  We also fit a similar model with the 1992-2019 basal area increment expressed as a percentage of 1992 basal area.  Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.  Variance explained for each model are reported using the approach described in Nakagaw et al. [-@Nakagawa2013], as implemented in the MuMIn package for R [@add later].

@@

#### Volume estimates
Tree volumes were estimated using species- and region-specific equations [@Nigh2016] based on tree height and DBH.  We estimated tree heights by fitting diameter-height models developed separately for fir and spruce, and for each period.  Trees with broken tops were omitted from diameter-height models.

@@

@@

# Fit model
htDiamFit<- 
  dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% # filter for live trees
    filter(is.na(DBT)) %>% # filter for trees with no broken tops
    drop_na(Height) %>% # remove trees where height was not measured
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create tree ID column
  lmer(Height~log(DBH)+(1|Plot/TreeID),data=.)

appNums("heightDiamRawPlot",display="FALSE")
```

Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiamRawPlot",display="cite")`).  A comparison of height-diameter models showed that intercepts and slopes of log-transformed heights did not vary significantly (p<0.05) between species or time periods (not shown), therefore a single height-diameter linear mixed-effects model was developed using `r dat %>% filter(Species %in% c("Bl","Sx")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset, with log-transformed diameter as the fixed effect, and tree nested within plot as random effect. This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log(DBH).  Residual plots were assessed to ensure goodness of fit.  This model was then applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@Nigh2016].

@@

@@

```{r plot height to diameter models by species and year, fig.width=6, fig.asp=1.2}

@@

>>>>

---
title: Analyses of 27 years of stand structure development and basal area growth response to a range of initial basal area density levels, 1992-2019 
author: "Mike Jull and Hardy Griesbauer"
date: "29/03/2020"
output:
  word_document:
    reference_docx: wordTemplate2.docx
  html_document:
    df_print: paged
csl: cjfr.csl
bibliography: ep1162Bib.bib
editor_options: 
  chunk_output_type: console
---

@@

# Load libraries
library(tidyverse)
library(here)
library(lme4)
library(lmerTest)
library(broom)
library(MuMIn)
library(ggeffects)
library(car)
library(flextable)
library(captioner)
library(lsmeans)

@@

# Some additional data prep

# set up diameter classes
diamClass=seq(from=7.5,to=95,by=5)
dat<- # assign variable
  dat %>% 
  mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F))

@@

### Tree-level growth responses
To test the hypothesis that tree-level growth response over the 1992-2019 period varied with interactions between (i) tree basal area at the time of treatment (1992); (ii) residual basal area; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 basal area increment as the response variable, and treatment, species and log-transformed 1992 basal area as interacting fixed effects, with plot as a random intercept effect.  We also fit a similar model with the 1992-2019 basal area increment expressed as a percentage of 1992 basal area.  Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.  Variance explained for each model are reported using the approach described in Nakagawa et al. [-@Nakagawa2013], as implemented in the MuMIn package for R [@add later].

@@

#### Volume estimates
Tree volumes were estimated using species- and region-specific equations [@nigh2016] based on tree height and DBH.  We estimated tree heights by fitting diameter-height models developed separately for fir and spruce, and for each period.  Trees with broken tops were omitted from diameter-height models.

@@

### Tree density
```{r sph by diam class species and time}

# need to create  a list of plots with five measurements
plotDenom<-
  dat %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
  group_by(BA.Target,Year,Plot) %>% 
  summarise(Num.Plots=n()) %>% 
  ungroup() %>% 
  dplyr::select(BA.Target:Plot) %>% 
  group_by(BA.Target,Year) %>% 
  summarise(Num.Plots=n())

# PLOTS
# Raw plot - examine SPH in each plot to detect where mortality occurred
# sphPlot.raw<-
#   dat %>% 
#   left_join(plotDat,by="Plot") %>% # join with plot data
#   filter(Status=="Live") %>% # filter for live trees only
#   group_by(BA.Target,Species,Plot,Year,DBH.class) %>% 
#   summarise(SPH=n()*20)  %>% 
#   filter(Species %in% c("Bl","Sx")) %>% # remove Fdi and Ep
#   ggplot()+
#   aes(x=DBH.class,y=SPH,group=Species,color=Species)+
#   geom_point()+
#   facet_grid(Plot~Year)+
#   theme(axis.text.x = element_text(angle = 90, hjust = 1),
#         legend.position="bottom")+
#   xlab("DBH Class (cm)")
# 
# appNums(name="sphRawPlot",display=FALSE)

# 1. SPH by diameter class and species for 1992 and 2019 periods
sphPlot1<-
  dat %>% 
  filter(Year %in% c(1992,2019)) %>% # Select for two time periods
    filter(Species %in% c("Bl","Sx")) %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
  group_by(BA.Target,Year,Species,DBH.class) %>% 
  summarise(SPH.Sum=n()*20) %>% # plot multiplier
  left_join(plotDenom,by=c("BA.Target","Year")) %>% 
  mutate(SPH=SPH.Sum/Num.Plots) %>%

  ggplot()+
  aes(x=DBH.class,y=SPH,group=Species,fill=Species,color=Species)+
  geom_point()+
  geom_path()+
    # geom_density(alpha=0.4)+
  facet_grid(Year~BA.Target)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
        legend.position = "bottom")+
  xlab("DBH Class (cm)")

  figNums(name="sphSpeciesPlot",display=FALSE)
```

```{r plot pct change in SPH by diam class between 92 and 19}

# % change in diameter class between 1992 and 2019 time periods by species
 sphPCTPlot<- 
  dat %>% 
  filter(Year %in% c(1992,2019)) %>% 
  filter(Status=="Live") %>% 
  filter(Species %in% c("Bl","Sx")) %>% 
  group_by(Plot,Species,Year,DBH.class) %>% 
  summarise(SPH.Total=n()) %>% 
  pivot_wider(names_from="Year",values_from="SPH.Total") %>% 
  mutate(Pct.Change=(`2019`-`1992`)/`1992`*100) %>% 
  mutate(Pct.Change=round(Pct.Change,3)) %>%
  drop_na(Pct.Change) %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data

  # Plotting
  ggplot()+
  aes(x=DBH.class,y=Pct.Change,fill=Species)+
  geom_boxplot()+
  geom_hline(yintercept=0, linetype="dashed", color="darkgrey")+
  facet_grid(Species~BA.Target)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
        legend.position="bottom")+
  xlab("DBH Class (cm)")+
  ylab("% change in SPH between 1992 and 2019")

  figNums(name="sphPCTPlot",display=FALSE)

# ggsave(filename="baDiamClass.png")

Mean plot tree density by diameter class diagrams showed that across all treatment levels, subalpine fir comprised the majority of stems in the smaller diameter classes, both in the year after treatment, and in 2019 (`r figNums("sphSpeciesPlot",display="cite")`). The diameter class distribution of fir in all treatment units was negative exponential.  Mean spruce density was lower than fir, with a relatively flatter diameter distribution.  Within the high-removal treatment unit, fir density increased mostly in the smaller diameter classes between 1992 and 2019; in the low-removal treatment unit, fir density increased most in trees between 12.5 to 27.5cm DBH; and in the control treatment unit, fir density increased mostly in trees above 22.5cm DBH ( `r figNums("sphPCTPlot",display="cite")`).  Spruce tree density increased less than fir, with the exception of an increase of over 250% in SPH in the smallest diameter class in the light-removal treatment unit.

## Stand-level changes

### Volume estimates

@@

# Fit model
htDiamFit<- 
  dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% # filter for live trees
    filter(is.na(DBT)) %>% # filter for trees with no broken tops
    drop_na(Height) %>% # remove trees where height was not measured
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create tree ID column
  lmer(log(Height)~log(DBH)+(1|Plot/TreeID),data=.)

# Apply fit model to data and create new height column
dat<-
  dat %>% 
  mutate(Height.Pred=coalesce(Height,exp(fixef(htDiamFit)[1]+fixef(htDiamFit)[2]*log(DBH))))

# Height model fit plot
htModelFit.plot<-
  dat %>% 
  ggplot(aes(x=Height,y=Height.Pred))+
  geom_point()+
  xlim(0,40)+
  ylim(0,40)+
  geom_smooth(method="lm",se=F)+
  ylab("Predicted height (m) from diameter-height model")+
  xlab("Measured height (m)")

# create captions
appNums("heightDiamRawPlot",display="FALSE")
appNums("htModelFit",display="FALSE")

Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiamRawPlot",display="cite")`).  A comparison of height-diameter models showed that intercepts and slopes of log-transformed heights did not vary significantly (p<0.05) between species or time periods (not shown), therefore a single height-diameter linear mixed-effects model was developed using `r dat %>% filter(Species %in% c("Bl","Sx")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset, with log-transformed diameter as the fixed effect, and tree nested within plot as random effect. This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log(DBH).  Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`appNums("htModelFit",display="cite")`).  The model slightly overestimated tree heights in shorter trees, and underestimated tree heights in taller trees.  This model was applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@Nigh2016].

```{r hardcode volume coefficients per Nigh}

volCoef<-

  ## TOTAL VOLUME
  #  Subalpine fir 
  data.frame(Species="Bl",VolType="total",b0=-9.79,b1=1.813,b2=1.033) %>%

  # Douglas-fir
  add_row(Species="Fd",VolType="total",b0=-10.19,b1=1.73,b2=1.206) %>%

  # Add spruce
  add_row(Species="Sx",VolType="total",b0=-10.024,b1= 1.801,b2= 1.107) %>%

  ## MERCHANTABLE VOLUME
  # Subalpine fir
  add_row(Species="Bl",VolType="merch",b0=-9.980,b1= 1.792,b2= 1.102) %>%

  # Douglas-fir
  add_row(Species="Fd",VolType="merch",b0=-10.503,b1= 1.724,b2= 1.296) %>%

  # Spruce
  add_row(Species="Sx",VolType="merch",b0=-10.280,b1= 1.769,b2= 1.206 ) %>%

  mutate(BGC.Zone="SBS") %>%

  # convert species to factor
  mutate(Species=factor(Species)) %>% 
  dplyr::select(Species,BGC.Zone,everything())

```{r generate volume estimates}

dat<-

  # Total volumes
  dat %>% 
  left_join(filter(volCoef,VolType=="total"),by="Species") %>% 
  mutate(VolTot=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
  dplyr::select(-c(BGC.Zone,VolType,b0,b1,b2)) %>%

  # Merchantable volume
  left_join(filter(volCoef,VolType=="merch"),by="Species") %>% 
  mutate(VolMerch=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
  dplyr::select(-c(BGC.Zone,VolType,b0,b1,b2)) %>%

  # basal area
  mutate(BA=(DBH/100/2)^2*pi)

# Create stand data
standDat<-
  dat %>% 
  group_by(Plot,Year) %>% 
  summarise(VolTot=sum(VolTot,na.rm=T)*20, # notice that all variables are by hectare
            VolMerch=sum(VolMerch,na.rm=T)*20,
            BA=sum(BA,na.rm=T)*20,
            k=n()*20) %>% 
  mutate(QMD=sqrt(BA/(k*0.0000785))) %>% 
  left_join(plotDat,by="Plot") %>%  # joinwith plot Data
   mutate(TU=paste("TU",BA.Target,sep=":"))  # create a TU column for plotting

```{r boxplot stand volumes by year}

volPlot<-
  standDat %>% 
  # plot
  ggplot()+
  aes(x=Year,y=BA,fill=BA.Target)+
  geom_boxplot()+
  facet_wrap(~TU)+
  theme(legend.position="none")+
  ylab("Mean plot volume (m/hectare)")

volFit<-
  standDat %>% 
  lm(VolTot~BA.Target*Year,data=.)

@@

```{r plot density by diameter class, fig.width=6}
sphPlot1

`r figNums(name="sphSpeciesPlot",caption="Mean density of stems per hectare (SPH) by diameter class and species, faceted by treatment unit and two measurement periods.  Lowest value is included in each diameter class, highest value is not.")`

```{r plot density change PCT by diameter class, fig.width=6}
sphPCTPlot

`r figNums(name="sphSpeciesPlot",caption="Percentage change between 1992 and 2019 in stems per hectare (SPH) by diameter class, faceted by treatment unit and species.   tsphPCTPlotwo measurement periods.  Lowest value is included in each diameter class, highest value is not.")`

@@

```{r plot height to diameter models by species and year, fig.asp=1.2, fig.width=6}

@@

##### page break
```{r plot diam to height model fit}

htModelFit.plot
```

`r appNums("htModelFit",caption="Predicted height against measured height for 870 tree measurements.  Outlier represents a tree with a broken top.")`
